Supermartingales and Termination Analysis of Statistical PCF

### 2020-12-26 LICS21/ranking.tex
Created version for LICS21 submission in subdirectory LICS21

\documentclass{article}

%% BEGIN {Luke's Macros}
%%
\newif\ifdraft
\drafttrue %% To hide all comments and highlighting, just comment this line out

\input{../Style/comment}
\usepackage{calculation}
\usepackage{graphicx}
\usepackage{tikz}
\usetikzlibrary{positioning, graphs, quotes, graphdrawing, shapes.geometric}
\usegdlibrary{layered, circular}
\usepackage[square]{natbib}
\setcitestyle{aysep={}}

\definecolor{oxblue}{RGB}{0,33,71}
\definecolor{oxgold}{HTML}{a0630a}
\usepackage[final,
bookmarks,
bookmarksopen,
colorlinks,
final,
linkcolor=red,
citecolor=oxgold,
pdfstartview=FitH ]{hyperref}

\usepackage{amsmath} % cleveref must be loaded afer amsmath
\usepackage{cleveref}
\Crefname{theorem}{Theorem}{Theorems}
\Crefname{corollary}{Corollary}{Corollary}
\crefname{proposition}{Proposition}{Propositions}
\Crefname{claim}{Claim}{Claims}
\Crefname{definition}{Definition}{Definitions}
\Crefname{fact}{Fact}{Facts}
\Crefname{conj}{Conjecture}{Conjectures}
\Crefname{example}{Example}{Examples}
\Crefname{remark}{Remark}{Remarks}
\Crefname{convention}{Convention}{Conventions}
\Crefname{lemma}{Lemma}{Lemmas}
\Crefname{assumption}{Assumption}{Assumptions}
\Crefname{section}{Section}{Sections}
\Crefname{appendix}{Appendix}{Appendices}
\Crefname{figure}{Figure}{Figures}
%\Crefname{equation}{Eq.}{Equations}

% \addtotheorempostheadhook[theorem]{\crefalias{thmlisti}{theorem}}
% \addtotheorempostheadhook[lemma]{\crefalias{thmlisti}{lemma}}
% \addtotheorempostheadhook[proposition]{\crefalias{thmlisti}{proposition}}
% \addtotheorempostheadhook[corollary]{\crefalias{thmlisti}{corollary}}
% \addtotheorempostheadhook[claim]{\crefalias{thmlisti}{claim}}
% \addtotheorempostheadhook[fact]{\crefalias{thmlisti}{fact}}
% \addtotheorempostheadhook[example]{\crefalias{thmlisti}{example}}
% \addtotheorempostheadhook[definition]{\crefalias{thmlisti}{definition}}
% \addtotheorempostheadhook[remark]{\crefalias{thmlisti}{remark}}
% \addtotheorempostheadhook[assumption]{\crefalias{thmlisti}{assumption}}

\newcommand{\Real}{\mathbb{R}}
\newcommand{\nnReal}{\mathbb{R}_{{\geq}0}}
\newcommand{\pReal}{\mathbb{R}_{{>}0}}

\newcommand\expect[1]{\mathbb{E}[#1]}
\newcommand\set[1]{\{#1\}}
\newcommand\dif{\mathrm{d}}
\newcommand\calF{\mathcal{F}}
\newcommand\calI{\mathcal{I}}
\newcommand\Leb{\mathrm{Leb}}
\newcommand\ndraw[2]{\#\mathrm{draw}_{#1}(#2)}

\makeatletter
\newcommand{\dotDelta}{{\vphantom{\Delta}\mathpalette\d@tD@lta\relax}}
\newcommand{\d@tD@lta}[2]{%
  \ooalign{\hidewidth$\m@th#1\mkern-1mu\cdot$\hidewidth\cr$\m@th#1\Delta$\cr}%
}

%\usepackage[bbgreekl]{mathbbol} %for \bblambda
%%
%% END {Luke's Macros}

\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{verbatim}
\usepackage{enumitem}
\newcommand{\tY}{\mathsf{Y}}
%\newcommand{\tif}[3]{\textsf{if }#1\textsf{ then }#2\textsf{ else }#3}
\newcommand{\tif}[3]{\mathsf{if}(#1, #2, #3)} % space-saving concrete syntax
\newcommand{\tsample}{\mathsf{sample}}
\newcommand{\tscore}{\mathsf{score}}
\newcommand{\skeletonPlaceholder}{\mathsf{X}} % I'm not sure what the best notation here is.
\DeclareMathOperator{\red}{red}
\DeclareMathOperator{\nnext}{next}
\DeclareMathOperator{\cbv}{cbv}

\usepackage{amsthm}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\newtheorem{example}{Example}

\theoremstyle{lemma}
\newtheorem{lemma}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{corollary}{Corollary}
\newtheorem{conjecture}{Conjecture}

\theoremstyle{remark}
\newtheorem*{remark}{Remark}

\title{Supermartingales and Termination Analysis of Statistical PCF}

\begin{document}

\maketitle

\lo{Tentative title}

\begin{abstract}
\lo{TODO}
\end{abstract}

\iffalse
\akr{Following your suggestion, I will be providing a criterion for termination of programs in PPCF \citep{DBLP:journals/pacmpl/EhrhardPT18} based on ranking supermartingales. 
As it's more convenient for this proof, a sampling-based semantics \lo{You need to provide reference(s) for sampling-based semantics.} will be used instead of the original distributional semantics. 
I assume some roughly applicable equivalence is proven somewhere, but it doesn't seem that hard anyway.}
\lo{Why not provide a proof as an appendix?}
\fi

\tableofcontents

Various theorems about probabilistic programs rely on the assumption that the program terminates almost surely. Proof rules based on relating the program state to supermartingales already exist for first-order imperative programs \citep{DBLP:journals/pacmpl/McIverMKK18}. This paper's contribution is to extend this method to a higher-order setting.

\section{Statistical PCF}

\subsection{Syntax of SPCF}

The language SPCF is a simply-typed lambda calculus with sampling of real numbers from $[0,1]$ and unbounded scoring, following \cite{MakOPW21}. Types and terms are defined as follows, where $r$ is a real number, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function, and $\Gamma$ is an environment:
\begin{align*}
  & \text{types } A, B ::= \textsf{R}  \mid  A \to B \\
  & \text{values } V ::= \lambda x.M  \mid  \underline{r} \\
  & \text{terms } M, N ::= V  \mid  x  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n)  \mid  \tY M  \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample  \mid  \tscore(M)
\end{align*}
\begin{align*}
  \frac{}{\Gamma ; x:A \vdash x:A} \qquad
  \frac{\Gamma ; x:A \vdash M : B}{\Gamma \vdash \lambda x.M : A \to B} \qquad
  \frac{}{\underline{r} : \textsf{R}} \qquad
  \frac{\Gamma \vdash M:A \to B \quad \Gamma \vdash N : A}{\Gamma \vdash M \, N : B} \\ \\
  \frac{\Gamma \vdash M_1:\textsf{R} \dots \Gamma \vdash M_n:\textsf{R}}{\Gamma \vdash \underline{f}(M_1,\dots,M_n) : \textsf{R}} \ f : \mathbb{R}^n \to \mathbb{R} \qquad
  \frac{\Gamma \vdash M : (A \to B) \to (A \to B)}{\Gamma \vdash \tY M : (A \to B)} \\ \\
  \frac{\Gamma \vdash M : \textsf{R} \quad \Gamma \vdash N_1 : A \quad \Gamma \vdash N_2 : A}{\Gamma \vdash \tif{M < 0}{N_1}{N_2} : A} \qquad
  \frac{}{\Gamma \vdash \tsample : \textsf{R}} \qquad
  \frac{\Gamma \vdash s : \textsf{R}}{\Gamma \vdash \tscore (s) : \textsf{R}}
\end{align*}

Terms are identified up to $\alpha$-equivalence, as usual. The set of all terms is denoted $\Lambda$, and the set of closed terms is denoted $\Lambda^0$.

\paragraph{}
To define the reduction relation, let \emph{evaluation contexts} be of the form:
\begin{align*}
  E ::= & \, [\,] \mid E \, M \mid V \, E \mid \underline{f}(\underline r_1,\dots ,\underline r_{k-1}, E, M_{k+1}, \dots, M_n) \\ & \mid \tY E \mid \tif{E<0}{M_1}{M_2} \mid \tscore (E)
\end{align*}
then a term reduces if it is formed by substituting a redex in a context i.e.
\begin{align*}
  E[(\lambda x.M) N] & \to E[M[N/x]] \\
  E[\underline f (\underline r_1, \dots , \underline r_n)] & \to E[\underline{f(r_1,\dots,r_n)}] \\
  E[\tY \lambda x. M] & \to E[\lambda z. M[(\tY \lambda x. M)/x] z] \text{ where $z$ is not free in $s$}\\
  E[\tif{\underline r < 0}{M_1}{M_2}] & \to E[M_1] \text{ where }r < 0 \\
  E[\tif{\underline r < 0}{M_1}{M_2}] & \to E[M_2] \text{ where }r \geq 0 \\
  E[\tsample] & \to E[\underline r] \text{ where } r \in [0,1] \\
  E[\tscore(\underline r)] & \to E[\underline r].
\end{align*}
\changed[lo]{We write $\to^\ast$ for the reflexive, transitive closure of $\to$.}
\lo{The $\tscore$ rule should have a side condition: if $r > 0$.}

Every closed well-typed term either is a value or reduces to another term.

\subsection{Sampling semantics}
\label{sec:sampling semantics}
This version of the reduction relation allows $\tsample$ to reduce to any number in $[0,1]$. To more precisely specify the probabilities, an additional argument is needed to determine the outcome of random samples. Let $ I = [0,1] \subset \mathbb{R} $, and let $S = I^{\mathbb{N}}$, with the \changed[lo]{Borel $\sigma$-algebra $\calF$} and the probability measure, denoted $\mu$, given by the limit of $1 \gets I \gets I^2 \gets \cdots$, where the maps are the projections that ignore the last element. Equivalently, a basis of measurable sets is $\prod_{i=0}^\infty X_i$ where $X_i$ are all Borel and all but finitely many are $I$, and $\mu (\prod_{i=0}^\infty X_i) = \prod_{i=0}^\infty \Leb(X_i)$.
The maps $\pi_h:S \to I, \; \pi_t:S \to S$ popping the first element are then measurable.
\changed[lo]{Following \cite{DBLP:conf/esop/CulpepperC17}, we call the probability space $(S, \calF, \mu)$ the \emph{entropy space}.}

The $\sigma$-algebra and measure on $\Lambda$ are defined by considering it as a disjoint union of equivalence classes under replacing all the real constants by a placeholder, where the measure on each class is that of $\mathbb{R}^n$, where $n$ is the number of real constants.
\changed[lo]{Precisely, following \cite{DBLP:conf/icfp/BorgstromLGS16},
we view $\Lambda$ as $\bigcup_{m\in\omega} \big(\mathsf{Sk}_m \times \Real^m \big)$,
where $\mathsf{Sk}_m$ is the set of SPCF terms with exactly $m$ place-holders (a.k.a.~\emph{skeleton terms}) for numerals.
Thus identified, we give $\Lambda$ the countable disjoint union topology of the product topology of the discrete topology on $\mathsf{Sk}_m$ and the standard topology on $\Real^m$.
Note that the connected components of $\Lambda$ have the form $\{M\} \times \Real^m$, with $M$ ranging over $\mathsf{Sk}_m$, and $m$ over $\omega$. 
We fix the Borel algebra of this topology to be the $\sigma$-algebra on $\Lambda$.}

The one-step reduction is given by the function $\red : \Lambda^0 \times S \to \Lambda^0 \times S$ where
\begin{equation}
\red(M,s) = \left\{
    \begin{array}{ll}
        (E[N],s) & \text{if } M = E[R], R \to N \text{ and } R \neq \tsample \\
        (E[\underline{\pi_h(s)}],\pi_t(s)) & \text{if } M = E[\tsample] \\
        (M,s) & \text{if } M \text{ is a value}
    \end{array} \right .
\end{equation}
\lo{Both $\to$ and $\red$ are called reduction relation.}

The result after $n$ steps is then simply $\red^n(M,s) = \overbrace{\red(...\red(}^n M,s)...)$, and the limit $\red^\infty$ can then be defined as a partial function as $\lim_{n \to \infty} \red^n(M,s)$ whenever that sequence becomes constant by reaching a value. A term $M$ terminates for a sample sequence $s$ if the limit $\red^\infty(M,s)$ is defined.

The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid M \text{ terminates at } s \}$ is measurable. 
We say that a term $M$ is \emph{almost-surely terminating} (AST) just if $\mu(\{s \mid M \text{ terminates at } s\}) = 1$.

\lo{Alternatively, define the \emph{runtime of $M$} to be the random variable 
\[
T_M(s) := 
\begin{cases}
\min \set{n \mid \pi_0(\red^n(M, s)) \textrm{ is a value}} & \hbox{if $\red^\infty(M,s)$ is defined}\\
\infty & \hbox{otherwise}
\end{cases}
\]
Equivalently, we say that $M$ is \emph{almost-surely terminating} (AST) if $T_M < \infty$ a.s.; 
and $M$ is \emph{positively almost-surely terminating} (PAST) if $\expect{T_M} < \infty$.}

For example, the term 
\[
(\tY \lambda f, n: \tif{\tsample - 0.5 < 0}{n}{f (n+1)} ) \, \underline{0},
\] 
which generates a geometric distribution, terminates on the set $S \setminus [0.5,1]^\mathbb N$, which has measure 1, therefore it terminates almost surely, whereas 
\[
\tif{\tsample - 0.5 < 0}{\underline 0}{(\tY \lambda x. x) \underline 0}
\] 
which terminates on the set $\pi_h^{-1}[[0,0.5]]$, has probability 0.5 of failing to terminate.

\paragraph{}
Note that the requirement that $M$ be closed is not important for the proofs of the theorems, but is simply required to ensure that $M$ reaches a value, rather than a general normal form like $x \, y$.

\paragraph{}
This definition of almost-sure termination is equivalent to that given in \citep{MakOPW21}, although the program semantics is stated in a slightly different way. In particular, the argument to $\tscore$ is not relevant to termination (except for the possibility that its argument's evaluation wouldn't terminate).

\iffalse
\lo{Your operational semantics does not maintain a record of the current weight of the reduction.
A.s.~termination does depend on $\tscore$: see \cite[\S 4.3]{MakOPW21}\footnote{\url{https://arxiv.org/abs/2004.03924}}.
I think it important to take the behaviour of $\tscore$ into account;
you should do it as a future task.}
\fi

\section{Supermartingales}
One approach to proving that a term terminates almost surely is to find some variant that is bounded below and, on average, decreases sufficiently quickly that it must eventually reach 0, similarly to the approach taken in 
%\citep{DBLP:journals/pacmpl/McIverMKK18} 
\cite{DBLP:conf/popl/FioritiH15} for imperative programs.

These variants are defined as functions from reachable terms (i.e.~possible states of the program's execution) to real numbers. Specifically, let the set of reachable terms from a given closed starting term $M$ be $Rch(M) := \{N \in \Lambda \mid M \to^* N \}$, with the $\sigma$-algebra induced as a subset of $\Lambda$.

\begin{definition}\rm
A \emph{ranking function on $M$} is a measurable function $f:\mathit{Rch}(M) \to \mathbb{R}$ such that $f(N) \geq 0$ for all $N$, and
\begin{enumerate}
    \item $f(E[\tY \lambda x. N]) \geq 1+ f(E[\lambda z. N[(\tY \lambda x. N)/x] z) \text{ where $z$ is not free in $N$}$
    \item $f(E[\tsample]) \geq \int_I f(E[\underline{x}]) \, \Leb(\mathrm{d}x)$

    \item $f(E[R]) \geq f(E[R'])$ for any other redex $R$, where $R \to R'$.
\end{enumerate}
We say that the ranking function $f$ is \emph{strict} if there exists $\epsilon > 0$ such that $f(N) = 0$ iff $N$ is a value, and for all $E$ and $R \to R'$, $f(E[R']) \leq f(E[R]) - \epsilon$.

Any closed term for which a ranking (respectively, strict ranking) function exists is called \emph{rankable} (respectively, \emph{strictly rankable}). 
\changed[lo]{For example, the term $(\tY \lambda x.x) \, \underline 0$ is not rankable.}
It will be demonstrated later that for any rankable term $M$, if $(M_n)_{n \in \omega}$ is the reduction sequence starting from $M$, then $(f(M_n))_{n\in\omega}$ is a supermartingale, and $M$ terminates almost surely, but first, some preliminaries about supermartingales.
\end{definition}

\input{ranking-supermartingales}

\section{Constructing Ranking Functions}
Although rankability implies almost-sure termination, the converse does not hold in general. For example,
\begin{equation}
\tif{{-}\tsample < 0}{\underline{0}}{(\tY \lambda x. x) \, \underline 0}
\label{ex:0 probability reachable}
\end{equation}
terminates in 3 steps with probability 1, but isn't rankable because $(\tY \lambda x. x) \, \underline 0$ is reachable, although that has probability 0. 
Not only is this counterexample AST, it's PAST. %positively almost surely terminating i.e.~the expected time to termination is finite.

@@

In fact, even the partial function $\tY \Theta N \mapsto 2$ alone is a partial ranking function for this term.
\section{Antitone ranking functions}

\begin{example}[Random walk]
\label{ex:ac-ranking}
\begin{enumerate}
\item 1D biased (towards 0) random walk
\[
M_1 = 
\big(\tY \, \lambda f n . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{3/2} f \, (n + 1)}\big) \, 10
\]

\item 1D unbiased random walk
\[
M_2 = 
\big(\tY \, \lambda f n . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}\big) \, 10
\]

\item 1D biased (away from 0) random walk
\[
M_3 = 
\big(\tY \, \lambda f n . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/3} f \, (n + 2)}\big) \, 10
\]

%While {x > 0} do {x := x-1 \oplus_2/3 x:= x+1}

%M2 = While {x > 0} do {x := x-1 \oplus_2/3 x:= x+2}
\end{enumerate}
\end{example}

The term $M_1$ is rankable, terminating in 31 $\tY$-steps on average, and $M_3$ only has a $1/1024$ chance of terminating, but in between, $M_2$ is AST, but isn't $\tY$-PAST, therefore it isn't rankable and \Cref{thm:rankable implies termination} is insufficient to prove its termination. We therefore want to find a generalised notion of ranking function so that $M_2$ becomes rankable, and then prove it sound i.e.~rankable in this generalised sense implies AST.

\paragraph{Idea.}

The definition of antitone-convex ranking function $f$ (say) for $M \in \Lambda^0$ is the same as that of ranking function except that in the case of $\tY$-redex, 
we require the existence of an antitone (meaning: $r < r'$ implies $\epsilon(r) \geq \epsilon(r')$) function $\epsilon : \nnReal \to \pReal$ such that the ranking function $f :\mathit{Rch}(M) \to \nnReal$ satisfies
\[
f(E[R']) \leq f(E[R]) - \epsilon(f(E[R])) 
\]
where $R \to R'$ is the $\tY$-redex rule.

\iffalse
\lo{An aside: This CBV $\tY$-rule seems cleaner:
\[
\big(\tY f^{A \to B} \, x^A \, . \, \theta^B \big) \, v \to \theta[(\tY f \, x \, . \, \theta) / f, v / x].
\]
We assume $\tY f^{A \to B} \, x^A \, . \, \theta^B$ is a value.}
\fi

\begin{definition}
We say that a supermartingale $(Y_n)_{n \in \omega}$ and a stopping time $T$ adapted to filtration $(\calF_n)_{n \in \omega}$ are \emph{antitone strict} if $Y_n \geq 0$, and there exists an antitone function $\epsilon : \nnReal \to \pReal$ satisfying
\[
\expect{Y_{n+1} \mid \calF_n} \leq Y_n - \epsilon (Y_n) \cdot 1_{\set{T > n}}
\]
\end{definition}

%\lo{The preceding definition does not place any constraint on the stopping time $T$ (other than that it is w.r.t.~the filtration $(\calF_n)_{n \in \omega}$). However for the definition to make sense, $T$ must be the stopping time $T(s) := \min \set{n \mid Y_n(s) = 0}$.}

\begin{theorem}
\label{thm:a-c strict}
Let $(Y_n)_{n \in \omega}$ be an antitone strict supermartingale via function $\epsilon : \nnReal \to \pReal$, and stopping time $T$. 
Then $T < \infty$ almost surely.
\end{theorem}

\begin{proof}
First, as $(Y_n)$ is a supermartingale, $\expect{Y_n} \leq \expect{Y_0}$. 
Therefore

\begin{calculation}
\expect{Y_n \mid T > n}
  \step[=]{rearranging terms}
\dfrac{\mathbb P[T > n] \, \expect{Y_n \mid T > n} 
+ 
\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n} 
- 
\mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
  \step[=]{definition of conditional expectation}
\dfrac{\expect{Y_n} - \mathbb P[T \leq n] \, \expect{Y_n \mid T \leq n}}{\mathbb P[T > n]}
  \step[\leq]{$Y_n \geq 0$ always}
\dfrac{\expect{Y_n}}{\mathbb P[T > n]}
  \step[\leq]{$(Y_n)_n$ is a supermartingale}
\dfrac{\expect{Y_0}}{\mathbb P[T > n]}
\end{calculation}

\emph{Claim}: For all $0 < x \leq 1$, 
\(
\mathbb P[T > B_x] \leq x,
\)
where $B_x = \left \lceil \dfrac{\expect{Y_0}+1}{x \; \epsilon(\expect{Y_0} \, x^{-1})} \right \rceil$.

\smallskip

As the convex hull of $\epsilon$ (the greatest convex function less than or equal to it) satisfies all the conditions assumed of $\epsilon$, in addition to being convex, assume wlog that $\epsilon$ is convex.

Assume for a contradiction that $\mathbb P[T > B_x] > x$.
Then, take $n \leq B_x$. 
We have
%\lo{Change of justification of the second equality below: event, not discrete r.v. Recall: expectation conditioning on r.v.~(resp.~event) is a r.v.~(resp.~number).}
\begin{calculation}
\expect{Y_n-Y_{n+1}}
  \step[=]{$\calF_n \subseteq \calF_{n+1}$, definition and linearity of conditional expectation}
\expect{Y_n - \expect{Y_{n+1} \mid \calF_n}}
  \step[\geq]{antitone strict assumption}
\expect{\epsilon(Y_n) \cdot 1_{\set{T > n}}}
%  \step[=]{definition of conditional expectation for a discrete variable}
\step[=]{definition of expectation conditioning on an \emph{event}}
\mathbb P[T > n]\ \expect{\epsilon(Y_n) \mid T > n}
  \step[\geq]{Jensen's inequality}
\mathbb P[T > n]\ \epsilon(\expect{Y_n \mid T > n})
  \step[\geq]{proved earlier}
\mathbb P[T > n]\ \epsilon\Big(\dfrac{\expect{Y_0}}{\mathbb P[T > n]}\Big)
  \step[>]{assumption, $\mathbb P[T > n] \geq \mathbb P[T > B_x] > x$}
x \; \epsilon\big(\expect{Y_0} \, x^{-1}\big).
\end{calculation}
Therefore, by a telescoping sum
\[
\expect{Y_{B_x}} = \expect{Y_{B_x}-Y_0+Y_0} \leq \expect{Y_0} - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1}) \leq -1 < 0 
\]
%\lo{$\expect{Y_0 - B_x \; x \; \epsilon(\expect{Y_0} \, x^{-1})}$}
which is a contradiction, therefore the claim must be true, therefore $P[T > n]$ tends to 0 as $n$ tends to infinity, therefore $T < \infty$ almost surely.
\end{proof}

Note that it is essential in this proof that $\epsilon$ be defined for all $\mathbb R_{\geq 0}$, not just the values that $(Y_n)_{n \in \omega}$ actually takes (or at least if $(Y_n)_{n \in \omega}$ is uniformly bounded, $\epsilon$ must be positive at the supremum as well as the actual values).

\begin{example}
\akr{Consider a random variable $X$, uniformly distributed in the interval $[-1,1]$, $T := \min \set{n \mid 2^{-n} < X}$, 
\[
Y_k := 
\begin{cases}
4-2^{-k} & \hbox{if $k < T$}\\
0 & \hbox{otherwise}
\end{cases} 
\]
} \lo{The following reformulation defines $T$ and $Y_k$ directly as a measurable function on a probability space (rather than via a r.v.~$X$).}

Take the probability space $(\Omega, \calF, \mathbb{P})$ where $\Omega$ is the closed interval of reals $[-1, 1]$, $\calF$ the Borel $\sigma$-algebra, and $\mathbb{P}$ the corresponding Lebesgue probability measure.
Let $\omega \in [-1, 1]$. Define random variables $T$ and $(Y_k)_{k \in \omega}$:
\begin{align*}
T(\omega) & := 
\begin{cases}
\min\set{n \in \mathbb{N} \mid \omega > 2^{-n}} & \hbox{if $\omega \in (0, 1]$}\\
\infty & \hbox{otherwise}
\end{cases} 
\\
Y_k(\omega) & := 
\begin{cases}
4-2^{-k} & \hbox{if $\omega \in [-1, 2^{-k}]$ (equivalently $k < T(\omega)$)}\\
0 & \hbox{otherwise}
\end{cases} 
\end{align*}
\lo{N.B.~$\set{T = n} = (\frac{1}{2^{n}}, \frac{1}{2^{n-1}}]$, and so, $\set{T > n} = [-1, \frac{1}{2^n}]$.}
Plainly, $T$ is a stopping time, and $(Y_n)_{n \in \omega}$ is a supermartingale, adapted to the filtration $(\calF_n)_{n \in \omega}$ where $\calF_n = \calF$ for all $n$. 
In this case, $(Y_n)_{n \in \omega}$ either tends to $4$ as $n \to \infty$, or drops to 0 at some point, with an exponentially decreasing probability. 
With respect to the antitone function $\epsilon(x) = \frac{4-x}{4}$ and stopping time $T$, we have that $(Y_n)_{n \in \omega}$ is an antitone strict supermartingale, except that $\epsilon(4) = 0$, and $T = \infty$ with probability $\frac 1 2$, even though $4$ is larger than any value $(Y_n)_{n \in \omega}$ can actually take, and $\epsilon(Y_n)$ never actually reaches $0$.
%\lo{@Andrew: I refer to the preceding highlighted claim. Did you mean: for all $n, s$, $\epsilon(Y_n(s)) > Y_n(s)$?} \akr{No. This phrasing should be clearer.}
\end{example}

\changed[lo]{Let $f$ be a ranking function on a closed SPCF term $M$.
%Recall that $(f(M_n))_{n \in \omega}$ is a supermartingale.
We say that $M$ is \emph{antitone rankable} by $f$ if the supermartingale $(f(M_n))_{n \in \omega}$ and the stopping time $T_M^{\tY}$, both adapted to the fitration $(\calF_n)_{n \in \omega}$ where $\calF_n := \sigma(M_1, \ldots, M_n)$, are antitone strict.}

\begin{theorem}[Antitone ranking function soundness] \label{thm:antitone rankable implies termination}
If a closed SPCF term $M$ is antitone rankable, then $T_M^{\tY} < \infty$ a.s.~(equivalently, $M$ is AST).
\end{theorem}
\begin{proof}
As in %Theorem \ref{thm:basicRankingSoundness}, 
\Cref{thm:rankable implies termination},
take the set $S$ of traces \lo{TODO: We should find a better name for traces $s \in S$.} as a probability space, and define the same random variables $T_n, X_n, Y_n, T_M^{\tY}$. As before, $(Y_n)_{n \in \omega}$ is a supermartingale, and because $M$ is now assumed to be antitone rankable, it is an antitone strict supermartingale with respect to the stopping time $T_M^{\tY}$. Therefore, it follows from \Cref{thm:a-c strict} that $T_M^{\tY}$ is almost-surely finite.
\end{proof}

\paragraph{}
As before, constructing antitone ranking fuctions completely is not necessary, and there is a corresponding notion of an antitone partial ranking function.

Define an \emph{antitone partial ranking function} on a closed term $M$ to be a partial function $F : Rch(M) \rightharpoonup \mathbb{R}$ such that for some antitone function $\epsilon : \mathbb{R}_{\geq 0} \to \mathbb{R}_{>0}$
\begin{itemize}
\item $f(N) \geq 0$ for all $N$ where $f$ is defined.
\item $f$ is defined at $M$.
\item For any $N$ in the domain of definition of $f$, evaluation of $N$ will eventually reach some $O$ which is either a value or in the domain of definition of $f$, and $f(N) \geq \mathbb E [f(O) + \epsilon(f(O)) \times \text{the number of $\tY$-reduction steps from $N$ to $O$}]$ (where $f(O)$ is taken to be 0 if $O$ is a value outside of the domain of $f$).
\end{itemize}

\begin{theorem}  \label{thm:antitone partial implies rankable}
  Every antitone partial ranking function is a restriction of an antitone ranking function.
\end{theorem}
\begin{proof}
  Take a closed term $M$ and an antitone partial ranking function $f$ on $M$, with a corresponding antitone function $\epsilon$. Assume wlog that $\epsilon$ is convex (as if it isn't, we can just take its convex hull instead). As in Theorem \ref{thm:partial implies rankable}, define $f_1 : Rch(M) \rightharpoonup \mathbb R$ by
  \begin{align*}
    f_1(N) &= f(N) \text{ whenever $f(N)$ is defined},\\
    f_1(V) &= 0 \text{ for values $V$ not in the domain of $f$.}
  \end{align*}

  Define $(\nnext(N,s),\_) = \red^n(N,s)$ for the least $n \geq 0$ such that it's in the domain of $f_1$, and $g(N,s) = \left | \{m < n \mid \red^m(N,s) \text{ is of the form } (E[\tY N'],s') \} \right |$. 
  The function $\nnext$ is well-defined (i.e.~$n$ is finite) for all $N \in Rch(M)$ by induction on the path from $M$ to $N$, by the third condition on antitone partial ranking functions. Define $f_2(N) = \int_S f_1(\nnext(N,s)) + \epsilon(f_1(\nnext(N,s))) g(N,s) \, \mu(\mathrm d s)$. For any term $N$ where $f$ is defined, $f_2(N) = f(N)$, and the value that $f_2$ would have at $N$ if $f$ were not defined at $N$ is $\leq f(N)$, by the third condition on antitone partial ranking functions. In order to show that $f_2$ is an antitone partial ranking function, it therefore suffices to show that the value that $f_2$ would have had at each term if $f$ were not defined at that term is at least the expectation of $f_2$ after one reduction step (plus $\epsilon(f_2(N))$ if the reduction step is a $tY$-reduction). For any term $N$ which is not of the form $E[R]$ for some $\tY$-redex $R$, this is trivial. If $R$ is a $\tY$-redex, then $\epsilon(f_2(N)) \leq \int_S \epsilon(f_1(\nnext(N,s)))\, \mu(\mathrm d s)$ by the convexity of $\epsilon$, because $n$ is bounded. Therefore, the (total) function $f_2$, which agrees with $f$ on $f$'s domain, is an antitone ranking function on $M$ (with the same function $\epsilon$ if it's convex).
\end{proof}

As a corollary, any term which admits an antitone partial ranking function terminates almost surely.

\paragraph{\Cref{ex:ac-ranking} revisited}
\akr{All of these ranking functions are slightly incorrect, because $\Xi \, \underline n$ reduces to something like $\Xi \, (\underline n + 1)$ then $(\lambda z. \text{something}[\Xi/f]) (\underline n + 1)$, rather than reducing the argument first. This could be fixed by appealing to the confluent semantics version of the antitone ranking theorem, but it would be nice to have at least some examples here illustrating how antitone ranking functions work, rather than actually saving all these examples until the end.

Even though the difference is quite trivial, this was actually the original motivating example for the confluent trace semantics.}

Programs $M_1$ and $M_2$ are AST. For $M_1$, the partial ranking function 
\[
\big(\tY \, \lambda f n . \, \tif{n=0}{0}{f \, (n-1) \oplus_{3/2} f \, (n+1)}\big) \, \underline x \mapsto 3x + 1
\] 
suffices to prove its termination (and could equivalently be considered an antitone partial ranking function with the constant antitone function $\epsilon_1(x) = 1$).

For $M_2$, define the function $g_1 : \nnReal \to \nnReal$ by $g_1(x) = \log(x+1) + 1$.
Using shorthand 
$\Theta_2 = \tY \, \lambda f n . \, 
\tif{n = 0}{0}{f \, (n - 1) \oplus_{1/2} f \, (n + 1)}$, 
we can define an antitone partial ranking function $f_2 : \mathit{Rch}(M_2) \rightharpoonup \nnReal$ 
%with antitone function $\epsilon_2$ defined as follows (for $n \in \mathbb N$):
defined as: for $n \in \mathbb N$
\begin{align*}
{\Theta_2} \, n 
&\mapsto 
g_2(n)
\\
0 &\mapsto 0.
\end{align*}
For $n \geq 1$, $\Theta_2 \, n$ reduces in several steps to either $\Theta_2 \, (n-1)$ or $\Theta_2 \, (n+1)$, each with probability $1/2$, with one $\tY$-reduction.
Now
\begin{align*}
g_2(n) - \frac{g_2(n-1) + g_2(n+1)} 2 
  &=  \log \Big(\frac{n+1}{\sqrt{n(n+2)}}\Big) \\
  &=  \frac 1 2 \log\left(\frac{(n+1)^2}{n(n+2)}\right) \\
  &=  \frac 1 2 \log\left(1 + \frac 1 {n(n+2)}\right) \\
  &>  \frac 1 {4n(n+2)} \\
  &>  \frac 1 2 \frac 1 {5(n+1)^2} + \frac 1 2 \frac 1 {5(n+3)^2}
\end{align*}
\changed[lo]{(The first inequality follows from the fact that for all $x \in (0, 1)$, $\log (1 + x) > x$.)}
Moreover $\Theta_2 \, 0$ reduces to $0$ with one $\tY$-reduction with a reduction in $g_2$ of $1$.
Therefore by setting $\epsilon_2(x) = \frac 1 {5(e^{x-1}+1)^2}$, the condition on how much $g_2$ must decrease is met. It is also defined at $M_2 = \Theta_2 \, 10$, and is non-negative, therefore it is an antitone partial ranking function, and $M_2$ is AST.

>>>>

\section{Constructing Ranking Functions}
\label{sec:ranking}

Although rankability implies almost-sure termination, the converse does not hold in general. For example,
\begin{equation}
\tif{{-}\tsample < 0}{\underline{0}}{(\tY \lambda x. x) \, \underline 0}
\label{ex:0 probability reachable}
\end{equation}
terminates in 3 steps with probability 1, but isn't rankable because $(\tY \lambda x. x) \, \underline 0$ is reachable, although that has probability 0. 
Not only is this counterexample AST, it's PAST. %positively almost surely terminating i.e.~the expected time to termination is finite.

@@

In fact, even the partial function $\tY \Theta N \mapsto 2$ alone is a partial ranking function for this term.
